---
title: "Advanced Statistical Techniques for Analyzing Ecological Monitoring Data"
author: "Kyle Rosenblad"
date: "2 April 2025"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: hide
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
# Load packages silently
library(unmarked)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(gridExtra)
library(viridis)
set.seed(9912)
```

# Executive Summary

This document illustrates how cutting-edge statistical methods can help us better understand where species live across a landscape, even when our surveys sometimes fail to detect them. With an approach called occupancy modeling, we can:

- **Account for imperfect detection** - Just because we don't see a species doesn't mean it is absent.
- **Identify important habitat features** that determine where a species occurs.
- **Create more accurate distribution maps** to inform conservation management.
- **Save time and money** by optimizing future monitoring schemes.

The methods shown here have been successfully applied to numerous biodiversity monitoring programs and endangered species assessments.

# The Challenge: Understanding Species Distributions with Imperfect Data

Conservation planners face a common challenge: **How do we know where a species truly lives?**

Survey data are valuable but can carry several important limitations:

- Species may go undetected even when they are present.

- Survey conditions may affect our ability to detect the species of interest.

- Limited resources mean we cannot monitor everywhere all the time, or even close.

```{r simulate_data, include=FALSE}
# Number of sites and surveys
n_sites <- 100
n_surveys <- 5

# Create a site-level covariate for occupancy (e.g., topographic complexity)
topo_comp <- rnorm(n_sites)

# Create a survey-level covariate for detection (e.g., weather conditions)
weather <- matrix(rnorm(n_sites * n_surveys), n_sites, n_surveys)

# Set parameter values for the linear predictor in the
# occupancy process
beta0 <- -0.2  # Intercept
beta1 <- 1.1   # Effect of topographic complexity on occupancy

# Generate occupancy probabilities for each site
psi <- plogis(beta0 + beta1 * topo_comp)

# Set parameter values for the linear predictor in the
# detection process
alpha0 <- -1.5   # Intercept; this value keeps detection probabilities fairly low
alpha1 <- -0.8  # Effect of weather variable on detection

# Set up an empty matrix, which we'll fill with simulated observation data
y <- matrix(NA, n_sites, n_surveys)

# Generate true occupancy states, which we'll call z.
# This model is what we might call a single-season occupancy model, in which we
# assume that the true occupancy state at each site does not change across repeat # surveys.
z <- rbinom(n_sites, 1, psi)

# Simulate detection data
for (i in 1:n_sites) {
  for (j in 1:n_surveys) {
    # If the site is occupied (z=1),
    # each survey could yield a detection
    # or a non-detection
    if (z[i] == 1) {
      # Compute detection probability
      p <- plogis(alpha0 + alpha1 * weather[i, j])
      # Simulate detection or non-detection using this probability
      y[i, j] <- rbinom(1, 1, p)
    } else {
      # Site is unoccupied (z=0), so species cannot be detected
      y[i, j] <- 0
    }
  }
}

# Create site data and detection data for visualization
site_data <- data.frame(
  Site = 1:n_sites,
  Topographic_Complexity = topo_comp
)

# Reshape detection data for visualization
detection_data <- data.frame(y)
names(detection_data) <- paste0("Survey_", 1:n_surveys)
detection_data$Site <- 1:n_sites
detection_long <- reshape2::melt(detection_data, id.vars = "Site", 
                                variable.name = "Survey", value.name = "Detected")
detection_long$Detected <- factor(detection_long$Detected, levels = c(0, 1), 
                                labels = c("Not Detected", "Detected"))

# Calculate naive occupancy
site_data$Detected_Ever <- apply(y, 1, function(x) any(x == 1))

# Format data for unmarked
site_covs <- data.frame(topo_comp = topo_comp)
obs_covs <- data.frame(
  weather = as.vector(t(weather))
)
umf <- unmarkedFrameOccu(y = y, siteCovs = site_covs, obsCovs = obs_covs)

# Fit models
m0 <- occu(~1 ~1, data = umf)  # Constant model
m1 <- occu(~1 ~topo_comp, data = umf)  # Topography-only model
m2 <- occu(~weather ~topo_comp, data = umf)  # Full model

# Model selection
models <- fitList(
  "ψ(.)p(.)" = m0,
  "ψ(topo_comp)p(.)" = m1,
  "ψ(topo_comp)p(weather)" = m2
)
ms <- modSel(models)

# Predict occupancy across topographic complexity gradient
topo_comp_seq <- seq(min(topo_comp), max(topo_comp), length.out = 100)
newdata <- data.frame(topo_comp = topo_comp_seq)
pred_psi <- predict(m2, type = "state", newdata = newdata)

# Predict detection across weather gradient
weather_seq <- seq(min(weather), max(weather), length.out = 100)
newdata_det <- data.frame(weather = weather_seq)
pred_p <- predict(m2, type = "det", newdata = newdata_det)

# Landscape prediction
grid_size <- 20
landscape <- expand.grid(x = 1:grid_size, y = 1:grid_size)
landscape$topo_comp <- with(landscape, 
                                sin(x/3) * cos(y/3) + rnorm(grid_size^2, 0, 0.3))
landscape_pred <- predict(m2, type = "state", 
                         newdata = data.frame(topo_comp = landscape$topo_comp))
landscape$occupancy <- landscape_pred$Predicted
```

## The Problem of Imperfect Detection

When we only look at raw survey data, we can **underestimate** where species truly occur because our surveys failed to detect it in certain locations:

```{r naive_vs_true, fig.height=5, fig.width=10}
# Compare naive and true occupancy
comparison <- data.frame(
  Metric = c("Sites where species was observed", "Sites where species actually occurs"),
  Percentage = c(mean(site_data$Detected_Ever) * 100, mean(z) * 100)
)

ggplot(comparison, aes(x = Metric, y = Percentage, fill = Metric)) +
  geom_col() +
  scale_fill_viridis_d(option = "D") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(title = "How Much We Miss: The Detection Gap",
       subtitle = "Comparing what we observe in surveys vs. actual species occurrence",
       y = "Percentage of Sites") +
  theme_bw() +
  theme(legend.position = "none",
        axis.title.x=element_blank())
```

The chart above, for an example species, shows the gap between what we observe during surveys and the actual species occurrence. This kind of discrepancy often arises because:

1. The species may be present but inactive during a survey.

2. Weather conditions or other factors may animals' behavior, and/or our ability to detect them.

3. Dense vegetation or other habitat features might hide the species.

4. We might simply not notice, correctly identify, or correctly record the species due to the imperfect nature of human observers. This also includes camera traps, AI, etc.--nobody is perfect!

# Our Approach: Accounting for Imperfect Detection

## How It Works

We use repeated surveys at the same locations over time to help us understand how often we miss species when they're actually present.

```{r detection_pattern, fig.height=4, fig.width=10}
# Create a small example dataset for a conceptual visual
set.seed(42)
example_sites <- 8
example_surveys <- 4
example_data <- matrix(0, nrow = example_sites, ncol = example_surveys)

# Known occupied sites with varying detection patterns
example_data[1, c(1, 3)] <- 1  # Detected in survey 1 and 3
example_data[2, 2] <- 1        # Detected only in survey 2
example_data[3, c(2, 3, 4)] <- 1  # Detected in 3 surveys
example_data[4, c(1, 4)] <- 1  # Detected in surveys 1 and 4
example_data[5, ] <- 0         # Never detected (may or may not be occupied)
example_data[6, ] <- 0         # Never detected (may or may not be occupied)
example_data[7, 4] <- 1        # Detected only in survey 4
example_data[8, c(1, 2, 3, 4)] <- 1  # Detected in all surveys

# Convert to long format for plotting
example_df <- data.frame(example_data)
names(example_df) <- paste0("Survey_", 1:example_surveys)
example_df$Site <- paste0("Site ", 1:example_sites)
example_long <- reshape2::melt(example_df, id.vars = "Site", 
                              variable.name = "Survey", value.name = "Detected")
example_long$Detected <- factor(example_long$Detected, 
                               levels = c(0, 1), 
                               labels = c("Not Detected", "Detected"))

# Plot the detection pattern
ggplot(example_long, aes(x = Survey, y = Site, fill = Detected)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_manual(values = c("Not Detected" = "#E5E5E5", "Detected" = "forestgreen")) +
  labs(title = "Species Detection Across Multiple Surveys",
       subtitle = "Different patterns help us understand detection probability") +
  scale_x_discrete(labels=c("Survey 1", "Survey 2", "Survey 3", "Survey 4"))+
  theme_minimal() +
  theme(axis.text.y = element_text(size = 11),
        axis.title=element_blank(),
        legend.title=element_blank())
```

By looking at these patterns, our statistical model can:

- Estimate the probability of detecting the species when it's present

- Adjust our occupancy estimates to account for sites where the species was present but never detected

- Consider the effects of habitat and survey conditions

## The Benefits of Our Approach

Our modeling approach provides several key advantages:

```{r benefits_table}
benefits <- data.frame(
  Benefit = c("More Accurate Distribution Maps", 
             "Identifies Key Habitat Features", 
             "Quantifies Survey Effectiveness",
             "Optimizes Future Survey Efforts"),
  Description = c("Accounts for locations where species was present but not detected",
                 "Reveals which environmental factors determine where species occur, such as temperature or tree cover",
                 "Clarifies effects of survey conditions on our ability to find the species of interest",
                 "Helps design better surveys to maximize detection while minimizing costs")
)

kable(benefits) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE)
```

# Results: What Drives This Species' Distribution and Probability of Detection?

## Habitat Relationships

Our analysis reveals that our species is much more likely to occur in high-quality habitat:

```{r habitat_relationship, fig.width=7, fig.height=5}
ggplot(data.frame(topo_comp = topo_comp_seq, 
                 psi = pred_psi$Predicted, 
                 lower = pred_psi$lower, 
                 upper = pred_psi$upper), 
       aes(x = topo_comp, y = psi)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, fill = "#5B9BD5") +
  geom_line(size = 1.2, color = "#2E75B6") +
  labs(x = "Habitat Quality Index", y = "Probability of Occurrence", 
       title = "How Habitat Quality Affects Species Occurrence",
       subtitle = "Shaded area shows confidence range") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

This chart shows that:

- As habitat quality increases, the chance that the species is present increases.

- In poor-quality habitat, there's less than a 25% chance of species occurrence.

- In high-quality habitat, there's more than a 75% chance of species occurrence.

- The blue shaded area shows our confidence around our model's estimates.

## Survey Effectiveness

Understanding how survey conditions affect our ability to detect the species:

```{r detection_relationship, fig.width=7, fig.height=5}
ggplot(data.frame(weather = weather_seq, 
                 p = pred_p$Predicted, 
                 lower = pred_p$lower, 
                 upper = pred_p$upper), 
       aes(x = weather, y = p)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, fill = "#ED7D31") +
  geom_line(size = 1.2, color = "#C55A11") +
  labs(x = "Weather Condition Index (lower = better)", y = "Probability of Detection", 
       title = "How Weather Affects Our Ability to Detect the Species",
       subtitle = "Species may be present but not detected under poor weather conditions") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

Key insights:

- Weather affects our ability to detect the species.

- Under optimal conditions, if the species is present, we have about an 80-90% chance of detecting it on average.

- Under the worst conditions, the chance of detecting the species, even when it is present, can drop to quite near zero. These would be wasteful days to survey!

- Repeat surveys are critical for understanding where the species is found, as well as how to optimize future monitoring strategies in light of factors that make detection difficult, like poor weather.

# Predicted Distribution Map

Based on our model, we can predict where the species is likely to occur across an entire landscape:

```{r landscape_prediction, fig.width=9, fig.height=6}
ggplot(landscape, aes(x = x, y = y, fill = occupancy)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Probability of\nOccurrence", option = "viridis", 
                      labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Predicted Species Distribution",
       subtitle = "Based on habitat relationships and accounting for imperfect detection") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

This map shows the estimated probability of the species occurring across the landscape, with:

- Yellow areas representing high probability of occurrence (>75%)
- Light blue and green areas representing moderate probability (25-75%)
- Dark blue areas representing low probability (<25%)

## Management Implications

This distribution map helps conservation managers to:

1. **Confidently prioritize conservation efforts** in areas with high occupancy probability
2. **Target future surveys** in areas with moderate probability to improve knowledge of where the species is found
3. **Identify the species' preferred habitats** that should be protected or restored
4. **Track changes over time** by repeating the analysis with new survey data and improved monitoring schemes

# Conclusions and Recommendations

Our analysis provides several key insights:

1. **The species' true distribution is more extensive** than raw survey data suggests
2. **Habitat quality is a strong driver** of where this species occurs
3. **Weather conditions substantially influence** our ability to detect the species
4. **Repeated surveys are essential** for accurate distribution assessment

## Next Steps

Based on these findings, we recommend:

1. **Prioritize protection** of high-quality habitat areas shown in dark blue
2. **Avoid surveying in poor weather** to maximize detection and avoid wasting resources
3. **Conduct at least 3-4 repeat visits** at each site for reliable results
4. **Expand survey efforts** to areas with moderate occupancy probability to improve our understanding of where this species does and does not occur

---

*This demonstration uses simulated data for a hypothetical species. For your specific project, we would adapt these methods to your target species and available data. Ideally, we can also help you design your data collection strategy before you begin.*
```